Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-09-20 manuscript/manuscript.tex
Finished the density-based discretization section
Significantly reduced this part. Used a Step 1, ..., approach to the
algorithm description, like I did for the adaptive discretization. Moved
the description of figure 2 to the caption and removed some redundancy.
Now that the algorithm is in steps, the "Algorithm summary" isn't needed
anymore.

The algorithm is comprised of the following steps:

\textit{Step 1}:
Define a linear function $\rho_l(r')$ that assumes
the same values as the normalised density $\rho_n(r')$ on the boundaries of the
tesseroid ($r_1$ and $r_2$):

@@

\noindent
in which $L_r^\text{orig}$ is the radial dimension of the original tesseroid and
$\delta$ is a positive constant henceforth called the \textit{delta ratio}.

on
the radius $R_\text{max}$ at which the maximum absolute difference takes
place:

\noindent where the delta ratio $\delta$ is a predefined constant value and
$L_r^\text{sm}/L_r$ is defined as the tesseroid size ratio, while
$L_r^\text{sm}$ and $L_r$ are the radial dimensions of the tesseroid that is
currently analysed and of the \emph{original} one, respectively.
On the first iteration, the former tesseroid corresponds to the
\emph{original} one, so $L_r^\text{sm}/L_r$ is equal to one, but for
future iterations this radial dimension will be computed as the
difference of the top and bottom radii of the subdividing tesseroid, making
$L_r^\text{sm}/L_r$ lower than one.
The tesseroid size ratio is intended to reduce the number of unneeded
discretizations and focus on the ones that will significantly reduce the
numerical error:
a big subdividing tesseroid with small maximum absolute difference would
cause more relative numerical error than a small subdividing tesseroid with
higher maximum absolute difference.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]
    {figures/density-based-discretization-algorithm.pdf}
\caption{
    Example intended to show how the density-based algorithm works.}
\label{fig:density-discretization-algorithm}
\end{figure*}

Once the \emph{original} tesseroid is discretized, we proceed to
subsequently apply the same algorithm on every subdividing tesseroid.
We will start from equation \ref{eq:density-reference-line} defining a
new linear function $\rho_l(r')$, an absolute difference with equation
\ref{eq:density-abs-diff} and evaluate the discretization inequation
\ref{eq:delta-density}, taking into account that now $R_1$ and $R_2$
are the radial boundaries of this smaller tesseroid.
If the inequation holds, the current tesseroid is discretized and the
algorithm is subsequently applied to the resulting subdividing tesseroids.

If on the contrary the inequation \ref{eq:delta-density} is not held,
then the tesseroid will not be discretized.
The same happens if the tesseroid radial dimension is bellow the
threshold of 1mm.
On the first case due to the fact that no more discretizations are
needed, and on the second one because smaller tesseroids will not
introduce meaningful increases in the accuracy of the computation.
This two conditions guarantee that the algorithm is finite.
Once any of them is met and no more discretizations are carried out,
every subdividing tesseroid is subjected to the modified adaptive
discretization algorithm and then a second order GLQ will be applied in
order to compute the gravitational field generated by it.

On Figure \ref{fig:density-discretization-algorithm}(a) we propose an
arbitrary normalised density function as an example of how this
algorithm works.
We plotted its corresponding linear function $\rho_l(r')$ and a
dashed line that represents the maximum absolute difference.
We will assume that the discretization inequation
\ref{eq:delta-density} holds in this case, so we must divide the
tesseroid at the depth at which the maximum difference occurs.
The dots on Figure \ref{fig:density-discretization-algorithm}
represent the discretization points, i.e. the radial boundaries of the
subdividing tesseroids produced by the algorithm.

On Figure \ref{fig:density-discretization-algorithm}(b) we apply the
algorithm on the shallower subdividing tesseroid.
In the same way, we plot the linear function and compute the maximum
absolute difference $\Delta \rho (r')$.
In this case, the tesseroid size-ratio $L_r^\text{sm}/L_r$ is lower
than one and will be taken into account in the discretization
inequation, which will be assumed to hold also in this case, leading to
the division of this tesseroid.

On Figure \ref{fig:density-discretization-algorithm}(c) we perform the
same analysis on the deeper subdividing tesseroid, and assuming that the
inequality holds in this case we also subdivide it.

Finally, if we suppose that the value of $\delta$ used in this example
does not produce further discretizations, the resulting subdividing
tesseroids can be seen on Figure
\ref{fig:density-discretization-algorithm}(d).

It's easy to see that the delta ratio indirectly controls how many
times the tesseroids will be subdivided based on the variation of the
density function.
Therefore it determines the accuracy involving the numerical error that
the density variation introduces and also the computation time: a lower
value of $\delta$ will produce more discretizations leading to an
increase in the accuracy of the approximation but also in the
computation time.
This raises the need to determine a maximum value of $\delta$ that
ensures an acceptable accuracy while minimising the computation time.

In a general way, the algorithm can be summarised in the following steps:

\begin{enumerate}
\renewcommand{\theenumi}{(\arabic{enumi})}
    \item Normalise the density function to the [0, 1] interval,
    \item Compute the absolute difference between the normalised
          density and a linear function that assumes the same values on the
          top and bottom radii that this normalised density.
    \item If product of the maximum of the absolute difference and a
          tesseroid size ratio is greater that a predefined delta ratio
          ($\delta$), then the tesseroid will be subdivided on the
          radius at which this maximum takes place. Otherwise, the
          tesseroid won't be split.
    \item If it's discretized, then apply items (2), (3) and (4) for
          every subdividing tesseroid.
\end{enumerate}

\subsection{Algorithm summary}

When computing any gravitational field generated by an arbitrary tesseroid on
a specific computation point, we present the following methodology:

\begin{enumerate}
    \renewcommand{\theenumi}{(\arabic{enumi})}
    \item Apply the density-based discretization algorithm in case of
          variable density producing subdivisions of the original tesseroid,
    \item Apply the modified adaptive discretization algorithm for
          each subdivision, generating a set of smaller tesseroids,
    \item Compute the desired gravitational field for every small tesseroid
          through the GLQ approximation with fixed quadrature orders
          (equation \ref{eq:glq-var-dens}).
\end{enumerate}

In most practical cases, the forward model will consist in a set of
tesseroids whose gravitational effect will be computed on a set of
computation points.
These steps can be applied on every tesseroid and for each computation point,
adding the contribution of each tesseroid to the resulting
gravitational field.

>>>>

The algorithm is comprised of the following steps
(Fig. \ref{fig:density-discretization-algorithm}):

\textit{Step 1}:
Define a linear function $\rho_l(r')$ that assumes
the same values as the normalised density $\rho_n(r')$ at the boundaries of the
tesseroid ($r_1$ and $r_2$):

@@

\noindent
in which $L_r$ is the radial dimension of the tesseroid being considered for division,
$L_r^\text{orig}$ is the radial dimension of the original tesseroid, and
$\delta$ is a positive constant henceforth called the \textit{delta ratio}.

\textit{Step 5}:
If inequality \ref{eq:delta-density} is not satisfied, then the tesseroid is split in
two parts at the radius $r_\text{max}$ at which the maximum absolute difference (Eq.
\ref{eq:density-abs-diff}) takes place.

\textit{Step 6}:
Repeat steps 1-6 for each smaller tesseroid produced in step 5.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]
    {figures/density-based-discretization-algorithm.pdf}
\caption{
    Example application of the density-based discretization algorithm to a non-linear
    density function.
    (a) the normalised density function $\rho_n(r')$ (blue), current boundaries of the
    tesseroid (orange dots), and the linear density function $\rho_l(r')$ (orange line).
    The dashed red line represents the maximum density difference $\Delta \rho (r')$ at
    which the tesseroid would be divided (assuming that the inequality
    \ref{eq:delta-density} is not satisfied).
    (b) second iteration of the algorithm with a new linear density function and maximum
    density difference. The tesseroid would be divided at the depth indicated by the
    dashed red line.
    (c) third iteration of the algorithm.
    (d) final output of the density-based discretization, assuming that all four new
    tesseroids satisfy inequality \ref{eq:delta-density}.
}
\label{fig:density-discretization-algorithm}
\end{figure*}

Once all smaller tesseroids satisfy Eq. \ref{eq:delta-density},
each one is subjected to the adaptive
discretization algorithm described earlier to calculate their gravitational effects.

On the first iteration, the ratio $L_r/L_r^\text{orig} = 1$ because the tesseroid being
divided is the original one.
For future iterations, the ratio will be progressively smaller than one as the
tesseroids get smaller.
This is intended to limit the number of divisions to the ones that will
significantly reduce the numerical error:
dividing a large tesseroid with a small $\text{max}\{ \Delta \rho(r') \}$ would
improve the integration accuracy more than dividing a small tesseroid with a
higher $\text{max}\{ \Delta \rho(r') \}$.
We also do not divide tesseroids with $L_r < 1\ mm$ because any integration errors would
be negligible due to the small mass of the tesseroid..

The higher $\delta$ is, the fewer divisions will be made, and vice-versa.
Thus, it controls how many times the tesseroids will be divided based on the density
function and, indirectly, determines the accuracy and computation time of
numerical integration.
This raises the need to determine a maximum value of $\delta$ that
ensures an acceptable accuracy while minimising the computation time.
